---
title: "Analysis_hke"
author: "Monica Gonzalez Castrillon"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: yes
  pdf_document:
    number_sections: no
    toc: yes
---


### Dissentangling climate and parental effects on recruitment of southern stock of European hake ###


In this study we examine the influence of environmental variables —sea surface temperature (SST), the North Atlantic Oscillation (NAO), and the 
Atlantic Multidecadal Oscillation (AMO)— alongside with spawning stock biomass (as a proxy for density dependence) and parental effects (size at first
maturity by sex and the length structure of the spawning stock) on recruitment dynamics of the southern hake stock in Iberian Atlantic waters over the last few decades. 
The annual variability in recruitment (R) was modelled using a Generalized Additive Model (GAM), with environmental and biological variables included as explanatory predictors. Additionally, we investigate the potential of the Shannon diversity index (H') as an indicator of the state of the southern hake population. 
The findings of this research will provide valuable insights into the drivers of recruitment variability in the southern hake stock, which are crucial for sustainable management under changing conditions and for evaluating the utility of tools to assess the stock's state.


```{r, include=FALSE}

# Load packages ----------------------------------------------------------------

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  PerformanceAnalytics, FSA, TSA, FSAdata, dplyr, tidyverse,
  magrittr, lattice, Publish, sciplot, multcomp, car,
  knitr, gplots, zoo, fpp2, tseries, imputeTS, forecast,
  lmtest, tidymv, ggplot2, GGally, Hmisc, fUnitRoots,
  itsadug, carData, corrplot, corrgram, DescTools, mgcv, gratia, patchwork
)

# Load data --------------------------------------------------------------------

model<-read.csv("modelo1.csv",header=TRUE,dec=".",sep=";")
model <- model %>% 
  mutate(across(c(ssb_val, ssb_low, ssb_upp, rec_value, rec_low, rec_upp,
                  L50_females), as.numeric)) %>%
  filter(year >= 1980) # Subset year

```


## Data exploration (Zuur et al 2010)

```{r, include=FALSE}

# Exploratory plots ------------------------------------------------------------

  # Scatterplots

vars_loess <- c("ssb_val", "L50_females", 
                "shannon_females_s1", "shannon_females_s2", 
                "shannon_females_s3", "shannon_females_s4", 
                "sst", "nao", "amo", "year")

for (var in vars_loess) {
  print(
    ggplot(model, aes_string(x = var, y = "rec_value")) +
      geom_point() +
      geom_smooth(method = "loess", se = TRUE) +
      labs(title = paste("Recruitment vs", var)) +
      theme_minimal()
  )
}

# Boxplots

box_vars <- c("rec_value", "ssb_val", "L50_females", 
              "shannon_females_s1", "shannon_females_s2", 
              "shannon_females_s3", "shannon_females_s4", 
              "sst", "nao", "amo", "year")

par(mfrow = c(2, 3)) 
for (v in box_vars) {
  boxplot(model[[v]], main = v)
}


# Outliers ---------------------------------------------------------------------
  
outlier_vars <- c("rec_value", "ssb_val", "L50_females","shannon_females_s1", 
                  "shannon_females_s2", "shannon_females_s3", "shannon_females_s4",
                  "sst", "nao", "amo"
)

model_clean <- model 

for (var in outlier_vars) {
  cat("Processing variable:", var, "\n")
  ts_data <- ts(model[[var]], start = 1980) 
  tso_res <- tsoutliers::tso(ts_data)

  if (!is.null(tso_res$outliers) && nrow(tso_res$outliers) > 0) {
    outlier_indices <- tso_res$outliers$ind
    outlier_years <- time(ts_data)[outlier_indices]
    
    cat("  Outliers detected in years:", outlier_years, "\n")
    
    model_clean[[var]][model_clean$year %in% outlier_years] <- NA
  } else {
    cat("  No outliers detected.\n")
  }
}

# Assumptions verification -----------------------------------------------------

# Normality

vars_to_check <- c("rec_value","ssb_val", "L50_females","shannon_females_s1", "shannon_females_s2",
                   "shannon_females_s3", "shannon_females_s4", "sst", "nao", "amo")

for (var in vars_to_check) {
  cat("\n\n---", var, "---\n")
  
  data <- model_clean[[var]]
  
  par(mfrow = c(1, 3)) 
  plot(data, main = paste("Plot of", var), ylab = var)
  hist(data, main = paste("Histogram of", var), xlab = var)
  qqnorm(data, main = paste("QQ Plot of", var))
  qqline(data, col = "red")
  par(mfrow = c(1, 1))
  
  if (length(na.omit(data)) >= 3 && length(na.omit(data)) <= 5000) {
    print(shapiro.test(data))
  } else {
    cat("Shapiro-Wilk test skipped (data size not in [3, 5000])\n") # Shapiro-Wilk Test
  }
}

# Homocedasticity 

predictors <- c("ssb_val", "L50_females","shannon_females_s1", "shannon_females_s2",
                "shannon_females_s3", "shannon_females_s4", "sst", "nao", "amo")

for (var in predictors) {
  formula <- as.formula(paste("rec_value ~", var))
  model_lm <- lm(formula, data = model_clean)
  
  plot(model_lm$fitted.values, model_lm$residuals,
       main = paste("Residuals vs Fitted for", var),
       xlab = "Fitted values", ylab = "Residuals", pch = 20, col = "darkblue")
  abline(h = 0, col = "red", lwd = 2)
  
  cat("\nBreusch-Pagan Test for", var, "\n")
  print(bptest(model_lm)) # Breusch-Pagan test result
}


# Correlation 

  # Corrplot
vars <- c("rec_value","ssb_val", "L50_females","shannon_females_s1", "shannon_females_s2",
                "shannon_females_s3", "shannon_females_s4", "sst", "nao", "amo")
numeric_vars<-model_clean[,vars]
cor_matrix<-cor(numeric_vars, method = "spearman", use="complete.obs")
round(cor_matrix,2)
par(mfrow=c(1,1))
corrplot::corrplot(cor_matrix,method="circle",tl.col="black") 

  # Pairwise plots
ggpairs(
  model_clean[,vars],
  lower=list(continuous=wrap("points",alpha=0.5)),  
  upper=list(continuous=wrap("cor",method="spearman")),
  diag=list(continuous="densityDiag")  
)

# Multicolineality 

vars_to_use <- vars[!vars %in% c("rec_value", "shannon_females_s2", "shannon_females_s4")]
full_formula <- as.formula(paste("rec_value ~", paste(vars_to_use, collapse = " + ")))
full_model <- lm(full_formula, data = model_clean)
vif_values <- vif(full_model)
print(vif_values)
high_vif <- vif_values[vif_values > 5]
cat("\nVariables with high multicollinearity (VIF > 5):\n") 
print(high_vif)

```

Due to high correlation (r > 0.60) between SST and AMO, these variables were 
excluded from the model. Similarly, because of high collinearity (VIF > 5) 
among the seasonal Shannon indices, seasons 2 and 4 were omitted from the model.


## GAM modelling 

Covariables in the model were selected with a forward stepwise procedure based 
on the Akaike information criterion (AIC) and an adjusted R-square. 
The best (and most parsimonious) model was finally chosen based on the 
compromise between low AIC values, a high adjusted R-square and significant predictors.

In this case, GAMs were fitted using cubic regression splines as the basis for 
the explanatory variables, with REML as the method for parameter estimation. 
The basis dimension was restricted to a k=4 to allow sufficient flexibility 
while avoiding overfitting problems. When a spline was not required (edf=1), 
the variable was considered a linear effect. 


```{r, message=FALSE, warning=FALSE, paged.print=FALSE}

# Variable selection and model fit ---------------------------------------------

m1<-gam(rec_value~ssb_val,data=model_clean)
m2<-gam(rec_value~ssb_val+L50_females,data=model_clean)
m3<-gam(rec_value~ssb_val+L50_females+s(shannon_females_s1,bs="cr",k=4),data=model_clean)
m4<-gam(rec_value~ssb_val+L50_females+s(shannon_females_s1,bs="cr",k=4)+
          s(shannon_females_s3,bs ="cr",k=4),data=model_clean) 
m5<-gam(rec_value~ssb_val+L50_females+s(shannon_females_s1,bs="cr",k=4)+
          s(shannon_females_s3,bs="cr",k=4)+s(nao,bs="cr",k=4),data=model_clean)
m6<-gam(rec_value~ssb_val+L50_females+s(shannon_females_s1,bs="cr",k=4)+
          s(shannon_females_s3,bs="cr",k=4)+s(nao,bs="cr",k=4),data=model_clean)


models <- list(m1, m2, m3, m4, m5, m6)
model_names <- paste0("m", 1:6)

  # Summaries of the model
for (i in seq_along(models)) {
  cat("\n--- Summary of model m", i, " ---\n", sep = "")
  print(summary(models[[i]]))
}

  # Comparison of models
results <- data.frame(
  Model = character(),
  AIC = numeric(),
  GCV = numeric(),
  sp = numeric(),
  r2 = numeric(),
  dev_expl = numeric(),
  stringsAsFactors = FALSE
)

for (i in seq_along(models)) {
  mod <- models[[i]]
  mod_summary <- summary(mod)
  results[i, ] <- list(
    Model = model_names[i],
    AIC = AIC(mod),
    GCV = mod$gcv.ubre,
    sp = mod_summary$sp.criterion,
    r2 = mod_summary$r.sq * 100,
    dev_expl = mod_summary$dev.expl * 100 
  )
}
print(results)

```

## Final GAM model 

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}

gam_final<-gam(rec_value~ssb_val+L50_females+s(shannon_females_s1,bs="cr",k=4)+
                 s(shannon_females_s3,bs="cr",k=4)+s(nao,bs="cr",k=4),data = model_clean)

summary(gam_final)
draw(gam_final)
#plot(gam_final, pages=1)
AIC(gam_final)
gam.check(gam_final) 

  # Model validation  

    # Residuales
plot(fitted(gam_final),resid(gam_final)) # Homocedasticity
abline(h=0,col="red")
acf(resid(gam_final),lag.max=38,main="ACF") # Autocorrelation function
pacf(resid(gam_final),lag.max =38,main="PACF") # Partial autocorrelation function

Box.test(resid(gam_final), type = "Ljung-Box") # Independence
#Ljung-Box test indicates a potential temporal trend
qqnorm(resid(gam_final))
qqline(resid(gam_final), col = "red")
shapiro.test(resid(gam_final)) # Normality assumption
jarque.bera.test(resid(gam_final))
adfTest(resid(gam_final),lag=1) # Stationary assumption
kpss.test(resid(gam_final),lshort=FALSE)
t.test(resid(gam_final),mu=0) # Zero mean assumption

# ARIMA for temporal dependecy
arma_res<-auto.arima(resid(gam_final),stationary=FALSE,trace=TRUE,seasonal=FALSE,
                     approximation=FALSE,stepwise=TRUE)
summary(arma_res)
checkresiduals(arma_res)

```
Because Ljung-Box test indicated a potential temporal trend, the auto.arima() function
was used to check for any temporal structure in the model residuals.
The best-fitting model was an ARMA (0,0,0) with zero mean
“The best-fitting time series model was an ARMA(0,0,0) with zero mean, 
indicating that the data showed no significant autocorrelation or moving 
average patterns. Essentially, the series behaves like white noise, 
with no detectable temporal dependence.


## Final Model plots

```{r}


# Partial effects

plots<-draw(gam_final,residuals=FALSE,rug=TRUE,draw=FALSE)
plots_titled<-Map(function(p,title) {
  p+
    theme_classic(base_size=12)+
    labs(title=title)+
    theme(axis.title=element_blank())
},plots,c("s(H´season 1)","s(H´season 3)","s(NAO)"))

final_plot<-wrap_plots(plots_titled,ncol=2)

final_plot<-final_plot+plot_layout(guides="collect")&
  theme_classic(base_size=12)&theme(axis.title.y=element_text(angle=90,vjust=1,size=14)
  )
print(final_plot)


# Increase in deviance explained plot

dev_exp<-c(summary(m2)$dev.expl,summary(m3)$dev.expl,summary(m4)$dev.expl,summary(gam_final)$dev.expl)

incremento<-c(NA,diff(dev_exp))
labels_incremento<-c("",paste0("\u2206",round(incremento[-1]* 100,1),"%"))
nombres_modelos<-c("SSB+L50","+H\u2032 s1","+H\u2032 s3","+NAO")

df_modelos<-data.frame(Modelo=factor(nombres_modelos,levels=nombres_modelos),
                       Deviance=dev_exp,Incremento=labels_incremento)

ggplot(df_modelos,aes(x=Modelo,y=Deviance))+
  geom_bar(stat="identity",fill="#C75B4A",alpha= 0.9)+
  geom_text(aes(label=Incremento),vjust=-0.5,size=4)+
  labs(y="Deviance Explained (%)",x="Stepwise Inclusion")+
  theme_classic(base_size=14) +
  theme(axis.text.x=element_text(face="italic")
  )


# Temporal tendency of all variables 

min_year<-min(model$year)
max_year<-max(model$year)
clean_x_axis<-theme(axis.title.x=element_blank(),axis.text.x=element_blank())

p1<-ggplot(model,aes(x=year,y=rec_value))+
  geom_ribbon(aes(ymin=rec_low,ymax=rec_upp),fill="lightblue",alpha=0.3)+
  geom_line(color="#08306B")+theme_classic()+labs(title="",y="R")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+
  clean_x_axis

p2<-ggplot(model,aes(x=year,y=ssb_val))+
  geom_ribbon(aes(ymin=ssb_low,ymax=ssb_upp),fill="lightblue",alpha=0.3)+
  geom_line(color="#5BAA9F")+theme_classic()+labs(y="SSB")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+
  clean_x_axis

p3<-ggplot(model,aes(x=year,y=L50_females))+
  geom_line(color="#A4C882")+theme_classic()+
  labs(y = expression(L[50]))+theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+
  clean_x_axis

p4<-ggplot(model,aes(x=year,y=nao))+
  geom_line(color="#C3B935")+theme_classic()+labs(x="",y="NAO")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))
  
p5<-ggplot(model,aes(x=year,y=shannon_females_s1))+
  geom_line(color="#D29B6E")+theme_classic()+labs(y="H´season 1")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+
  scale_y_continuous(labels=scales::number_format(accuracy=0.1))+ 
  clean_x_axis

p6<-ggplot(model,aes(x=year,y=shannon_females_s2))+
  geom_line(color="#8B2E25")+theme_classic()+labs(y="H´season 2")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+
  clean_x_axis

p7<-ggplot(model,aes(x=year,y=shannon_females_s3))+
  geom_line(color="#A97C91")+theme_classic()+
  labs(x="",y="H´season 3")+theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks=seq(min_year,max_year,by=5))+ 
  clean_x_axis

p8<-ggplot(model,aes(x=year,y=shannon_females_s4))+
  geom_line(color="#E7B6B7")+theme_classic()+labs(x="",y="H´season 4")+
  theme(axis.title.y=element_text(size=9))+
  scale_x_continuous(limits=c(min_year,max_year),breaks= seq(min_year,max_year,by=5))

p1+p2+p3+p4+plot_layout(ncol=1)
p5+p6+p7+p8+plot_layout(ncol=1)

```
